---
title: "Credit Risk Scoring: End-to-End ML with skiPatrol"
subtitle: "Feature Store → tidymodels → Model Registry → SPCS Deployment"
format: html
---

An end-to-end credit risk scoring workflow demonstrating how a fintech company
can build, evaluate, and deploy credit-risk models entirely within Snowflake
using R and the `skiPatrol` package (with `skiLift` for DBI/dbplyr
connectivity).

**Runs unmodified in three environments** -- local, Posit Native App, and
Workspace Notebooks -- via `sfr_connect()`/`dbConnect(Snowflake())`'s
auto-detection. See `RUNBOOK.md` and `_config.R`. Validated locally
end-to-end 7 Sep 2026 against a real account; supersedes
`01_demo_samiul_original.qmd`, which used ODBC and a hand-built Snowpark
session as Native-App-specific workarounds -- see `RUNBOOK.md` for why
neither is needed with the P1-02 fixes.

**What you'll do:**

1. Explore lending data already in Snowflake
2. Build reusable features in the Snowflake Feature Store
3. Train competing models with tidymodels (logistic regression vs XGBoost)
4. Evaluate and compare model performance
5. Register both models in the Snowflake Model Registry with metrics
6. Select the champion version programmatically and deploy via SPCS
7. Score new applications via REST and SQL
8. Verify autocapture (ML Observability)
9. *(Optional)* Set up drift/performance Model Monitoring

**Database layout** (created by `00_setup.qmd`):

| Schema | Contents |
|--------|----------|
| `RAW_DATA` | Source tables — customers, payments, loans |
| `FEATURES` | Feature Store (entity, feature views, dynamic tables) |
| `TRAINING` | Materialised training datasets |
| `MODELS` | Model Registry (versioned models, metrics, services) |

## 1. Setup and Connect

```{r}
#| label: setup
library(skiPatrol)
library(skiLift)
library(tidymodels)
library(xgboost)
library(ggplot2)
library(dplyr)
library(scales)
source("_config.R")
cat("tidymodels", as.character(packageVersion("tidymodels")), "loaded\n")
cat("xgboost", as.character(packageVersion("xgboost")), "loaded\n")
```

### Connect and Configure Database Context

No hand-built Snowpark session and no ODBC: `sfr_connect()` auto-detects the
right strategy for whichever environment this runs in (local
`connections.toml`, Native App OAuth profile, or a Workspace Notebook
session -- see `_config.R` and `RUNBOOK.md`). Database/schema/warehouse are
still passed explicitly, regardless of environment: several registry/feature
store calls (e.g. `sfr_write_table()`) issue unqualified SQL that relies on
the session's *current* database, which auto-detection alone doesn't always
set.

```{r}
#| label: connect
conn <- sfr_connect(
  name = CR_CONNECTION_NAME, database = CR_DATABASE,
  schema = CR_REGISTRY_SCHEMA,
  warehouse = if (!is.null(CR_WAREHOUSE) && nzchar(CR_WAREHOUSE)) CR_WAREHOUSE else NULL
)

# skiPatrol's sfr_connection has no DBI/dbplyr methods -- it's a Python/ML
# session wrapper. DBI/dbplyr connectivity is skiLift's job (per skiPatrol's
# own DESCRIPTION), so a separate skiLift connection backs the dbplyr table
# refs below. This replaces the ODBC dbi_con Samiul's Native-App version used.
dbi_con <- DBI::dbConnect(
  Snowflake(), name = CR_CONNECTION_NAME, database = CR_DATABASE,
  schema = CR_SOURCE_SCHEMA, warehouse = CR_WAREHOUSE
)

tbl_customers <- fqn_source("CREDIT_CUSTOMERS")
tbl_payments  <- fqn_source("CREDIT_PAYMENTS")
tbl_loans     <- fqn_source("CREDIT_LOANS")

n <- sfr_query(conn, paste("SELECT COUNT(*) AS N FROM", tbl_loans))
cat(sprintf("Connected. Loan applications: %s\n", format(n$N, big.mark = ",")))
```

### Model Registry

```{r}
#| label: registry
reg <- sfr_model_registry(conn,
  database = CR_DATABASE,
  schema   = CR_REGISTRY_SCHEMA
)
cat(sprintf("Model Registry: %s.%s\n", CR_DATABASE, CR_REGISTRY_SCHEMA))
```

## 2. Feature Store

Define reusable features in the **FEATURES** schema using `dplyr`/`dbplyr`
pipelines. Each Feature View is backed by a Snowflake Dynamic Table that
refreshes automatically.

Features defined here are used identically in training and production scoring —
**define once, use everywhere**.

```{r}
#| label: feature-store-init
fs <- sfr_feature_store(conn,
  database  = CR_DATABASE,
  schema    = CR_FEATURE_SCHEMA,
  warehouse = conn$warehouse,
  create    = TRUE
)

# dbi_con (skiLift) and tbl_customers/tbl_payments/tbl_loans were set up in
# the connect chunk above -- reused here, not re-created.
customers_tbl <- tbl(dbi_con, I(tbl_customers))
payments_tbl  <- tbl(dbi_con, I(tbl_payments))
loans_tbl     <- tbl(dbi_con, I(tbl_loans))
cat("Feature Store and dbplyr table references ready.\n")
```

### Create Entity

```{r}
#| label: entity
customer_entity <- sfr_create_entity(
  fs, "CUSTOMER", "CUSTOMER_ID",
  desc = "Loan applicant"
)
customer_entity
```

### Feature View 1: Customer Profile

Static demographic features. We build the feature query as a `dbplyr` pipeline,
render to SQL with `dbplyr::sql_render()`, and pass that SQL string to
`sfr_create_feature_view()`.

```{r}
#| label: fv-profile
profile_query <- customers_tbl |>
  select(CUSTOMER_ID, AGE, ANNUAL_INCOME, EMPLOYMENT_LENGTH_YEARS,
         HOME_OWNERSHIP) |>
  mutate(LOG_INCOME = log(pmax(ANNUAL_INCOME, 1)))

profile_sql <- dbplyr::sql_render(profile_query)
cat("dbplyr pipeline rendered to SQL:\n")
cat(as.character(profile_sql), "\n")
```

```{r}
#| label: fv-profile-register
fv_profile <- sfr_create_feature_view(
  fs,
  name      = "CUSTOMER_PROFILE_FV",
  version   = "v1",
  entities  = customer_entity,
  feature_df  = profile_sql,
  desc      = "Customer demographic profile features",
  overwrite = TRUE
)
cat("Registered CUSTOMER_PROFILE_FV (using SQL string)\n")
```

### Feature View 2: Payment Behaviour (Dynamic Table)

Aggregated payment history — captures delinquency patterns. Passes the
`dbplyr` lazy table directly (contrast with the SQL string approach above).

Uses `refresh_freq` to create a **Dynamic Table** that Snowflake refreshes
automatically — use this when source data changes frequently.

```{r}
#| label: fv-payments
payment_query <- payments_tbl |>
  group_by(CUSTOMER_ID) |>
  summarise(
    AVG_DAYS_PAST_DUE  = mean(DAYS_PAST_DUE, na.rm = TRUE),
    MAX_DAYS_PAST_DUE  = max(DAYS_PAST_DUE, na.rm = TRUE),
    PCT_MISSED         = sum(ifelse(PAYMENT_STATUS == "MISSED", 1, 0),
                             na.rm = TRUE) / n(),
    PCT_ON_TIME        = sum(ifelse(PAYMENT_STATUS == "ON_TIME", 1, 0),
                             na.rm = TRUE) / n(),
    AVG_PAYMENT_AMOUNT = mean(PAYMENT_AMOUNT, na.rm = TRUE),
    TOTAL_PAYMENTS     = n(),
    .groups = "drop"
  )

fv_payments <- sfr_create_feature_view(
  fs,
  name         = "PAYMENT_BEHAVIOR_FV",
  version      = "v1",
  entities     = customer_entity,
  feature_df   = payment_query,
  refresh_freq = "1 hour",
  desc         = "Payment behaviour aggregates (dynamic table)",
  overwrite    = TRUE
)
cat("Registered PAYMENT_BEHAVIOR_FV (dynamic table, 1h refresh)\n")
```

### Feature View 3: Credit History

Aggregated lending history. Also passes the `dbplyr` lazy table directly.

```{r}
#| label: fv-credit
credit_query <- loans_tbl |>
  group_by(CUSTOMER_ID) |>
  summarise(
    TOTAL_APPLICATIONS = n(),
    AVG_LOAN_AMOUNT    = mean(LOAN_AMOUNT, na.rm = TRUE),
    MAX_LOAN_AMOUNT    = max(LOAN_AMOUNT, na.rm = TRUE),
    PRIOR_DEFAULTS     = sum(DEFAULTED, na.rm = TRUE),
    AVG_INTEREST_RATE  = mean(INTEREST_RATE, na.rm = TRUE),
    .groups = "drop"
  )

fv_credit <- sfr_create_feature_view(
  fs,
  name      = "CREDIT_HISTORY_FV",
  version   = "v1",
  entities  = customer_entity,
  feature_df  = credit_query,
  desc      = "Credit / lending history aggregates",
  overwrite = TRUE
)
cat("Registered CREDIT_HISTORY_FV\n")
```

```{r}
#| label: list-fvs
cat("Feature Views registered:\n")
sfr_list_feature_views(fs)
```

## 3. Exploratory Data Analysis

```{r}
#| label: eda-data
eda_query <- loans_tbl |>
  left_join(
    customers_tbl |>
      select(CUSTOMER_ID, AGE, ANNUAL_INCOME,
             EMPLOYMENT_LENGTH_YEARS, HOME_OWNERSHIP),
    by = "CUSTOMER_ID"
  )

loans_df <- collect(eda_query)
loans_df$DEFAULTED <- factor(loans_df$DEFAULTED, levels = c(0, 1),
                             labels = c("No Default", "Default"))
cat(sprintf("Joined dataset: %d rows x %d cols\n", nrow(loans_df), ncol(loans_df)))
```

### Default Rate Overview

```{r}
#| label: fig-default-dist
#| fig-cap: "Loan default distribution (~15% default rate)"
ggplot(loans_df, aes(x = DEFAULTED, fill = DEFAULTED)) +
  geom_bar(show.legend = FALSE) +
  geom_text(stat = "count", aes(label = comma(after_stat(count))),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Loan Default Distribution",
       subtitle = "~15% default rate -- realistic class imbalance",
       x = NULL, y = "Applications") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

### Income vs Default

```{r}
#| label: fig-income
#| fig-cap: "Lower income correlates with higher default probability"
ggplot(loans_df, aes(x = ANNUAL_INCOME, fill = DEFAULTED)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(labels = dollar) +
  coord_cartesian(xlim = c(0, 250000)) +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Income Distribution by Default Status",
       x = "Annual Income", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")
```

### Employment Length vs Default Rate

```{r}
#| label: fig-employment
#| fig-cap: "Longer employment correlates with lower default rate"
emp_summary <- loans_df |>
  mutate(emp_bin = cut(EMPLOYMENT_LENGTH_YEARS,
                       breaks = c(-1, 2, 5, 10, 20, 50),
                       labels = c("0-2yr", "3-5yr", "6-10yr",
                                  "11-20yr", "20+yr"))) |>
  group_by(emp_bin) |>
  summarise(default_rate = mean(DEFAULTED == "Default"),
            n = n(), .groups = "drop")

ggplot(emp_summary, aes(x = emp_bin, y = default_rate)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = sprintf("%.0f%%", default_rate * 100)),
            vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Default Rate by Employment Length",
       x = "Employment Length", y = "Default Rate") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## 4. Generate Training Data from Feature Store

Join the label spine (loan applications with default outcome) with all three
Feature Views. The Feature Store handles the joins automatically — the same
features used here will be retrieved at inference time.

```{r}
#| label: training-data
training_data <- sfr_generate_training_data(
  fs,
  spine = paste(
    "SELECT application_id, customer_id, defaulted",
    "FROM", tbl_loans
  ),
  features = list(fv_profile, fv_payments, fv_credit),
  spine_label_cols = "DEFAULTED"
)

tbl_training <- fqn_training("CREDIT_RISK_TRAINING")
sfr_write_table(conn, tbl_training, training_data, overwrite = TRUE)

cat(sprintf("Training data: %d rows x %d cols -> %s\n",
            nrow(training_data), ncol(training_data), tbl_training))
head(as.data.frame(training_data), 5)
```

```{r}
#| label: prep-labels
training_data$DEFAULTED <- factor(training_data$DEFAULTED,
                                  levels = c(1, 0),
                                  labels = c("yes", "no"))

cat("Label distribution:\n")
table(training_data$DEFAULTED)
```

## 5. Model Training with tidymodels

Two models compared with cross-validation:

- **Logistic Regression** — interpretable baseline, standard in credit risk
- **XGBoost** — gradient-boosted trees, typically better predictive performance

### Preprocessing Recipe

```{r}
#| label: recipe
credit_recipe <- recipe(
  DEFAULTED ~ AGE + ANNUAL_INCOME + LOG_INCOME +
    EMPLOYMENT_LENGTH_YEARS + HOME_OWNERSHIP +
    AVG_DAYS_PAST_DUE + MAX_DAYS_PAST_DUE +
    PCT_MISSED + PCT_ON_TIME + AVG_PAYMENT_AMOUNT +
    TOTAL_PAYMENTS + TOTAL_APPLICATIONS + AVG_LOAN_AMOUNT +
    MAX_LOAN_AMOUNT + PRIOR_DEFAULTS + AVG_INTEREST_RATE,
  data = training_data
) |>
  step_dummy(HOME_OWNERSHIP) |>
  step_impute_median(all_numeric_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

credit_recipe
```

### Define Model Specifications

```{r}
#| label: specs
logistic_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

xgb_spec <- boost_tree(
  trees = 500,
  tree_depth = 6,
  learn_rate = 0.05,
  min_n = 10
) |>
  set_engine("xgboost") |>
  set_mode("classification")

cat("Model specifications defined\n")
```

### Create Workflows

```{r}
#| label: workflows
wf_logistic <- workflow() |>
  add_recipe(credit_recipe) |>
  add_model(logistic_spec)

wf_xgb <- workflow() |>
  add_recipe(credit_recipe) |>
  add_model(xgb_spec)

cat("Workflows created\n")
```

### Cross-Validated Evaluation

5-fold stratified CV for honest performance estimates.

```{r}
#| label: cv
set.seed(42)
folds <- vfold_cv(training_data, v = 5, strata = DEFAULTED)
credit_metrics <- metric_set(roc_auc, accuracy, pr_auc)

cat("Evaluating logistic regression (5-fold CV)...\n")
results_lr <- fit_resamples(wf_logistic, folds, metrics = credit_metrics)

cat("Evaluating XGBoost (5-fold CV)...\n")
results_xgb <- fit_resamples(wf_xgb, folds, metrics = credit_metrics)

cat("Cross-validation complete\n")
```

## 6. Model Evaluation

### Compare CV Metrics

```{r}
#| label: cv-comparison
metrics_lr  <- collect_metrics(results_lr) |> mutate(model = "Logistic Regression")
metrics_xgb <- collect_metrics(results_xgb) |> mutate(model = "XGBoost")

comparison <- bind_rows(metrics_lr, metrics_xgb) |>
  select(model, .metric, mean, std_err) |>
  arrange(.metric, desc(mean))

as.data.frame(comparison)
```

```{r}
#| label: fig-cv-compare
#| fig-cap: "Model comparison: 5-fold cross-validation"
ggplot(comparison, aes(x = .metric, y = mean, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                position = position_dodge(width = 0.7), width = 0.2) +
  geom_text(aes(label = sprintf("%.3f", mean)),
            position = position_dodge(width = 0.7), vjust = -0.8, size = 3.5) +
  scale_fill_manual(values = c("#3498db", "#e67e22")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Model Comparison: 5-Fold Cross-Validation",
       x = "Metric", y = "Mean Score", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")
```

### Train Final Models and ROC Curves

Fit both models on the full training set and compare ROC curves on a holdout.

```{r}
#| label: final-fit
set.seed(42)
split <- initial_split(training_data, prop = 0.8, strata = DEFAULTED)
train_set <- training(split)
test_set  <- testing(split)

fit_lr  <- fit(wf_logistic, data = train_set)
fit_xgb <- fit(wf_xgb, data = train_set)

preds_lr  <- augment(fit_lr, test_set)
preds_xgb <- augment(fit_xgb, test_set)

roc_lr  <- roc_curve(preds_lr, DEFAULTED, .pred_yes) |> mutate(model = "Logistic")
roc_xgb <- roc_curve(preds_xgb, DEFAULTED, .pred_yes) |> mutate(model = "XGBoost")

holdout_lr <- bind_rows(
  roc_auc(preds_lr, DEFAULTED, .pred_yes),
  pr_auc(preds_lr, DEFAULTED, .pred_yes),
  accuracy(preds_lr, DEFAULTED, .pred_class)
) |> mutate(model = "Logistic Regression")

holdout_xgb <- bind_rows(
  roc_auc(preds_xgb, DEFAULTED, .pred_yes),
  pr_auc(preds_xgb, DEFAULTED, .pred_yes),
  accuracy(preds_xgb, DEFAULTED, .pred_class)
) |> mutate(model = "XGBoost")

cat("Holdout metrics:\n")
bind_rows(holdout_lr, holdout_xgb) |>
  select(model, .metric, .estimate) |>
  as.data.frame()
```

```{r}
#| label: fig-roc
#| fig-cap: "ROC curves: Logistic Regression vs XGBoost"
roc_data <- bind_rows(roc_lr, roc_xgb)

auc_lr  <- roc_auc(preds_lr, DEFAULTED, .pred_yes)$.estimate
auc_xgb <- roc_auc(preds_xgb, DEFAULTED, .pred_yes)$.estimate

ggplot(roc_data, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1.1) +
  geom_abline(linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("#3498db", "#e67e22"),
    labels = c(sprintf("Logistic (AUC=%.3f)", auc_lr),
               sprintf("XGBoost (AUC=%.3f)", auc_xgb))) +
  labs(title = "ROC Curves: Logistic Regression vs XGBoost",
       x = "False Positive Rate", y = "True Positive Rate", color = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = c(0.7, 0.25))
```

### Select Best Model Locally

```{r}
#| label: best-model
if (auc_xgb >= auc_lr) {
  best_fit  <- fit_xgb
  best_name <- "XGBoost"
  best_auc  <- auc_xgb
} else {
  best_fit  <- fit_lr
  best_name <- "Logistic Regression"
  best_auc  <- auc_lr
}
cat(sprintf("Best model: %s (AUC = %.3f)\n", best_name, best_auc))
```

## 7. Register Models and Select Champion

Register **both** candidate models as versions of `SFR_CREDIT_RISK_SCORER` in
the Model Registry, attach holdout metrics, then programmatically select the
champion based on registry-stored metrics.

The registry is the single source of truth for model governance — not local
R variables.

```{r}
#| label: verify-local
sfr_predict_local(best_fit, test_set[1:5, ])
```

```{r}
#| label: register-models
input_cols <- sfr_input_cols(
  training_data,
  exclude = c("DEFAULTED", "CUSTOMER_ID", "APPLICATION_ID")
)
cat(sprintf("Inferred %d input columns from training data\n", length(input_cols)))

extra_deps <- c("r-xgboost", "r-ranger", "numpy<2.0")

# task= is required for the optional Model Monitoring section (## 11) --
# sfr_add_monitor() rejects a version logged without it ("Registry model
# must be logged with task in order to be monitored"). Harmless to always
# set: pure registry metadata, no effect on registration/deploy otherwise
# (confirmed live 7 Sep).
mv_lr <- sfr_log_model(
  reg,
  model        = fit_lr,
  model_name   = "SFR_CREDIT_RISK_SCORER",
  predict_pkgs = c("tidymodels"),
  input_cols   = input_cols,
  output_cols  = list(prediction = "string"),
  conda_deps   = extra_deps,
  task         = "TABULAR_BINARY_CLASSIFICATION",
  comment      = sprintf("Logistic regression -- AUC %.3f", auc_lr)
)

mv_xgb <- sfr_log_model(
  reg,
  model        = fit_xgb,
  model_name   = "SFR_CREDIT_RISK_SCORER",
  predict_pkgs = c("tidymodels", "xgboost"),
  input_cols   = input_cols,
  output_cols  = list(prediction = "string"),
  conda_deps   = extra_deps,
  task         = "TABULAR_BINARY_CLASSIFICATION",
  comment      = sprintf("XGBoost -- AUC %.3f", auc_xgb)
)

cat("Both models registered\n")
```

### Attach Metrics

```{r}
#| label: attach-metrics
for (ver in list(list(mv = mv_lr,  h = holdout_lr),
                 list(mv = mv_xgb, h = holdout_xgb))) {
  for (i in seq_len(nrow(ver$h))) {
    sfr_set_model_metric(reg, "SFR_CREDIT_RISK_SCORER",
                         ver$mv$version_name,
                         ver$h$.metric[i], ver$h$.estimate[i])
  }
  cat(sprintf("  %s: attached %d metrics\n",
              ver$mv$version_name, nrow(ver$h)))
}
cat("All metrics stored in Model Registry\n")
```

### Compare Versions from Registry

```{r}
#| label: registry-compare
versions <- sfr_show_model_versions(reg, "SFR_CREDIT_RISK_SCORER")

metrics_list <- Filter(Negate(is.null), lapply(versions$name, function(vn) {
  m <- sfr_show_model_metrics(reg, "SFR_CREDIT_RISK_SCORER", vn)
  if (length(m) == 0 || is.null(m$roc_auc)) return(NULL)
  data.frame(version = vn,
             roc_auc  = as.numeric(m$roc_auc),
             pr_auc   = as.numeric(m$pr_auc),
             accuracy = as.numeric(m$accuracy))
}))

registry_comparison <- do.call(rbind, metrics_list)
cat("Model versions and metrics from registry:\n")
registry_comparison
```

### Select Champion Version

Pick the version with the highest `roc_auc` and set it as the model's
default version — the version downstream consumers use automatically.

```{r}
#| label: set-champion
champion_idx <- which.max(registry_comparison$roc_auc)
champion_ver <- registry_comparison$version[champion_idx]

sfr_set_default_model_version(reg, "SFR_CREDIT_RISK_SCORER", champion_ver)

cat(sprintf("Champion: %s (roc_auc = %.3f, pr_auc = %.3f, accuracy = %.3f)\n",
            champion_ver,
            registry_comparison$roc_auc[champion_idx],
            registry_comparison$pr_auc[champion_idx],
            registry_comparison$accuracy[champion_idx]))
cat("Set as default version in Model Registry\n")
```

## 8. Deploy Champion and Inference

Deploy the champion as a scalable REST service via Snowpark Container Services
with **autocapture** enabled. Autocapture logs every inference request/response
to the model's inference table automatically.

::: {.callout-warning}
**Deploying a fresh R/conda container takes 7+ minutes** (confirmed locally
7 Sep -- `sfr_deploy_model()` blocks synchronously for the whole image
build + push + service start; not a Native-App-specific delay). **Never run
this chunk live for the first time in front of an audience** -- deploy well
ahead of the session (see `RUNBOOK.md`'s pre-flight checklist). This chunk
checks whether the service is already `READY` and skips redeploying if so,
so re-running the notebook during rehearsal doesn't cost another 7 minutes;
set `FORCE_REDEPLOY <- TRUE` to force a fresh deploy anyway.
:::

```{r}
#| label: deploy
FORCE_REDEPLOY <- FALSE

svc_fqn <- paste(CR_DATABASE, CR_REGISTRY_SCHEMA, "CREDIT_RISK_SVC", sep = ".")
service_ready <- tryCatch({
  identical(sfr_get_service_status(reg, "credit_risk_svc")$status, "READY")
}, error = function(e) FALSE)

if (service_ready && !FORCE_REDEPLOY) {
  cat("Service already READY -- skipping redeploy (set FORCE_REDEPLOY <- TRUE to force).\n")
} else {
  deploy_start <- Sys.time()
  sfr_create_compute_pool(conn, "R_CREDIT_POOL",
                          instance_family = "CPU_X64_M")
  sfr_create_image_repo(conn, fqn_model("R_CREDIT_IMAGES"))
  sfr_execute(conn, paste("DROP SERVICE IF EXISTS", svc_fqn))

  sfr_deploy_model(
    reg,
    model_name   = "SFR_CREDIT_RISK_SCORER",
    version_name = champion_ver,
    service_name = "credit_risk_svc",
    compute_pool = "R_CREDIT_POOL",
    image_repo   = fqn_model("R_CREDIT_IMAGES"),
    force        = TRUE,
    autocapture  = TRUE
  )
  cat(sprintf("sfr_deploy_model() returned after %.1f min\n",
              as.numeric(difftime(Sys.time(), deploy_start, units = "mins"))))
}
```

```{r}
#| label: wait-service
sfr_wait_for_service(reg, "credit_risk_svc",
                     timeout_min = 10, poll_sec = 15)
```

### Score New Applications

Retrieve features from the Feature Store for a batch of customers and score
them through the deployed model.

```{r}
#| label: score-rest
scoring_features <- sfr_retrieve_features(
  fs,
  spine = paste(
    "SELECT DISTINCT customer_id FROM", tbl_customers,
    "LIMIT 20"
  ),
  features = list(fv_profile, fv_payments, fv_credit)
)

cat(sprintf("Retrieved %d rows x %d cols from Feature Store\n",
            nrow(scoring_features), ncol(scoring_features)))
```

```{r}
#| label: predict-rest
predictions <- sfr_predict(
  reg, "SFR_CREDIT_RISK_SCORER",
  scoring_features[, names(input_cols)],
  version_name = champion_ver,
  service_name = "credit_risk_svc"
)

results <- cbind(
  scoring_features[, "CUSTOMER_ID", drop = FALSE],
  predictions
)
cat("Predictions for 20 customers:\n")
results
```

### Score from SQL

The SPCS service exposes a `PREDICT` function callable directly from SQL —
accessible to any Snowflake user, BI tool, or scheduled task without R/Python.

```{r}
#| label: predict-sql
svc_fn   <- paste(CR_DATABASE, CR_REGISTRY_SCHEMA,
                  "CREDIT_RISK_SVC!PREDICT", sep = ".")
col_list <- paste(names(input_cols), collapse = ", ")

sql_score <- paste(
  "SELECT *,", svc_fn, "(", col_list, ") AS prediction",
  "FROM", fqn_training("CREDIT_RISK_TRAINING"),
  "LIMIT 10"
)

cat("SQL prediction query:\n")
cat(sql_score, "\n\n")
sfr_query(conn, sql_score)
```

### Verify Autocapture

Every inference request was logged to the model's inference table (ML
Observability). This feeds drift detection, performance tracking, and
monitoring.

::: {.callout-note}
**Expected delay, not a bug — do not wait on this live.** Autocapture writes to a
Snowflake-managed inference table through an asynchronous pipeline, the same
architecture SPCS uses for event-table logs/metrics — it is not synchronous with the
inference response. Snowflake doesn't publish a hard SLA, but the architecture (and
event-table precedent) points to minutes-scale latency, roughly 1–10 minutes in
practice. "Immediate visibility" in Snowflake's own framing means *you don't have to
build your own capture pipeline*, not *zero lag*. Failed requests are never captured at
all, only successful ones. This fully explains local testing seeing 0 rows after 60s
(7 Sep) — that's within the expected window, not a defect.

**For a live demo:** run this chunk once, non-blocking. If rows are already there
(e.g. because scoring happened a few minutes ago earlier in the session, or during
pre-flight), show them. If not, say so plainly and move on — don't poll live for
minutes waiting for it to appear.
:::

```{r}
#| label: autocapture
model_fqn <- paste(CR_DATABASE, CR_REGISTRY_SCHEMA,
                   "SFR_CREDIT_RISK_SCORER", sep = ".")

row_count <- sfr_query(
  conn,
  sprintf("SELECT COUNT(*) AS N FROM TABLE(INFERENCE_TABLE('%s'))", model_fqn)
)$N[1]

if (is.na(row_count) || row_count == 0) {
  cat("No autocapture rows visible yet -- expected, asynchronous pipeline (see note above).\n")
  cat("Check again in a few minutes rather than waiting here.\n")
} else {
  cat(sprintf("Autocapture rows captured: %d\n", row_count))
}
```

```{r}
#| label: autocapture-detail
#| eval: false
detail_sql <- sprintf("
SELECT
    TIMESTAMP,
    RESOURCE_ATTRIBUTES:\"snow.model.version.name\"::STRING AS MODEL_VERSION,
    RECORD_ATTRIBUTES:\"snow.model_serving.function.name\"::STRING AS FUNCTION_NAME,
    RECORD_ATTRIBUTES:\"snow.model_serving.response.code\"::INT AS RESPONSE_CODE
FROM TABLE(INFERENCE_TABLE('%s'))
ORDER BY TIMESTAMP DESC
LIMIT 10
", model_fqn)

cat("Recent autocapture entries:\n")
sfr_query(conn, detail_sql)
```

## 9. Optional: Model Monitoring

::: {.callout-note}
**Optional extra** — skip this section for a shorter demo; it doesn't affect anything
before it. Sets up drift/performance monitoring on the deployed model via
`skiPatrol`'s Model Monitoring API (`R/monitoring.R`), backed by Snowflake's
`MODEL_MONITOR_*` SQL functions. See `snowdropML/vignettes/model-monitoring.Rmd` for
the general API and Snowflake's own Model Registry/Monitoring docs for the underlying
concepts.

**Validated live, 7 Sep** -- three real constraints not obvious from the vignette
(found by testing against the real account, not documented anywhere beforehand):

1. **The model version must be logged with `task =`** (added to Section 7 above) --
   monitoring a version registered without it fails outright
   ("Registry model must be logged with task in order to be monitored").
2. **Classification models need `prediction_class_columns`/`actual_class_columns`**,
   not `prediction_score_columns`/`actual_score_columns` (score columns are for
   regression). **`sfr_monitor_source()` didn't expose the class-column parameters
   at all before this session** -- added as part of validating this section (see
   `RUNBOOK.md`).
3. **Class columns must be numeric-encoded (`0`/`1`), not string labels** -- our
   deployed model returns `"yes"`/`"no"`, which gets encoded below. The timestamp
   column must be `TIMESTAMP_NTZ`/`DATE`, not `TIMESTAMP_LTZ` (what
   `CURRENT_TIMESTAMP()` returns by default) -- also handled below.

**Stats and performance metrics compute immediately** -- confirmed live, no scheduled-
refresh delay, unlike autocapture's asynchronous pipeline in Section 8. **Drift metrics are not
included here** -- they require a baseline distribution to compare against
(`sfr_add_monitor()`'s source config accepts a `baseline` table, not yet exposed via
`sfr_monitor_source()`), which is unset by default and needs its own setup. Flagged as
a follow-on, not blocking.
:::

```{r}
#| label: monitoring-setup
monitor_name <- "CREDIT_RISK_MONITOR"
log_tbl <- fqn_model("INFERENCE_LOG")

existing_monitor <- tryCatch({
  sfr_get_monitor(reg, name = monitor_name)
  TRUE
}, error = function(e) FALSE)

if (existing_monitor) {
  cat(sprintf("Monitor %s already exists -- skipping setup.\n", monitor_name))
} else {
  # Stand-in "recent inference traffic": score CREDIT_RISK_TRAINING rows again,
  # spread over the last 3 days, with the real known outcome as ACTUAL. A real
  # deployment would source this from genuinely new scored requests (e.g. the
  # autocapture inference table from Section 8, once its timing is confirmed --
  # see RUNBOOK.md), not synthetic re-scoring of training data.
  svc_fn_call <- paste0(svc_fn, "(", col_list, ")")
  sfr_execute(conn, paste(
    "CREATE OR REPLACE TABLE", log_tbl, "AS",
    "SELECT",
    "  APPLICATION_ID,",
    "  DATEADD('hour', -UNIFORM(0, 72, RANDOM()), CURRENT_TIMESTAMP())::TIMESTAMP_NTZ AS EVENT_TIME,",
    "  IFF(", svc_fn_call, ":prediction::STRING = 'yes', 1, 0) AS PREDICTION,",
    "  DEFAULTED AS ACTUAL",
    "FROM", fqn_training("CREDIT_RISK_TRAINING"),
    "LIMIT 500"
  ))
  cat(sprintf("Inference log ready: %s\n", log_tbl))

  monitor_source <- sfr_monitor_source(
    log_tbl,
    timestamp_column = "EVENT_TIME",
    prediction_class_columns = "PREDICTION",
    actual_class_columns = "ACTUAL",
    id_columns = "APPLICATION_ID"
  )
  monitor_config <- sfr_monitor_config(
    model_name = "SFR_CREDIT_RISK_SCORER",
    version_name = champion_ver,
    warehouse = conn$warehouse,
    aggregation_window = "1 day"
  )
  sfr_add_monitor(reg, monitor_name = monitor_name, monitor_source, monitor_config)
  cat(sprintf("Monitor %s created.\n", monitor_name))
}
```

```{r}
#| label: monitoring-read
cat("Daily prediction volume:\n")
print(sfr_monitor_stats(conn, monitor_name, "COUNT", "PREDICTION",
                         start_time = Sys.Date() - 7, end_time = Sys.Date() + 1))

cat("\nDaily classification accuracy:\n")
print(sfr_monitor_performance(conn, monitor_name, "CLASSIFICATION_ACCURACY",
                               start_time = Sys.Date() - 7, end_time = Sys.Date() + 1))

cat("\nDaily F1 score:\n")
print(sfr_monitor_performance(conn, monitor_name, "F1_SCORE",
                               start_time = Sys.Date() - 7, end_time = Sys.Date() + 1))
```

## 11. Optional: SQL Pushdown with orbital

If the `orbital` package is installed, the trained tidymodels workflow can be
converted to pure SQL — useful for batch scoring without deploying an SPCS
service.

```{r}
#| label: orbital
#| eval: false
# pak::pak("tidymodels/orbital")
# library(orbital)
# sql_model <- orbital(fit_xgb)
# predict(sql_model, con = dbi_con, table = tbl_training)
```

## 12. Cleanup (Optional)

```{r}
#| label: cleanup
#| eval: false
# Delete the monitor first -- it runs on an hourly refresh against a real
# warehouse until deleted/suspended (confirmed live 7 Sep: refresh_interval
# defaults to "1 hour" regardless of aggregation_window).
# sfr_delete_monitor(reg, "CREDIT_RISK_MONITOR")
# sfr_execute(conn, paste("DROP TABLE IF EXISTS", fqn_model("INFERENCE_LOG")))

# Drop SPCS services before deleting the model
# sfr_execute(conn, paste("DROP SERVICE IF EXISTS", svc_fqn))
# sfr_execute(conn, paste("DROP COMPUTE POOL IF EXISTS R_CREDIT_POOL"))

# Drop model
# sfr_execute(conn, paste(
#   "DROP MODEL IF EXISTS",
#   paste(CR_DATABASE, CR_REGISTRY_SCHEMA, "SFR_CREDIT_RISK_SCORER", sep = ".")
# ))

# Full teardown (careful!)
# sfr_execute(conn, paste("DROP DATABASE IF EXISTS", CR_DATABASE))
```

## Summary

| Schema | What we did |
|--------|-------------|
| `RAW_DATA` | Queried source lending data |
| `FEATURES` | Created entity + 3 Feature Views (1 Dynamic Table) |
| `TRAINING` | Generated training data from Feature Store joins |
| `MODELS` | Registered both models, compared metrics, deployed champion |

### Key Advantages Over Traditional Workflow

| Traditional (RStudio + Plumber) | Snowflake (skiPatrol + SPCS) |
|---------------------------------|-------------------------------|
| Data exported to local machine | Data stays in Snowflake |
| Manual feature engineering scripts | Feature Store: define once, use everywhere |
| Model files on shared drives | Model Registry: versioned, governed, with metrics |
| Plumber API on a VM | SPCS: auto-scaled containers, no infra management |
| No inference logging | Autocapture: every prediction logged automatically |
| Manual model comparison | Registry metrics: programmatic champion selection |
